\plaintitle

### 2017-01-08 drafts/labor/piecework_lit.tex
piecework lit update

\topic{The HCI community has used the term ``piecework'' to describe
myriad instantiations of on--demand labor, but
researchers have generally made this allusion in passing.}
Since this paper traces a much stronger parallel between
(historical) piecework and (contemporary) on--demand work
(and, more generally, on--demand labor),
a more comprehensive background on piecework will be useful.
We will more carefully discuss piecework in this section
to help make our observations and arguments with better familiarity with the topic.
Specifically,
first, we'll define ``piecework'' as researchers in its field understand it;
and second, we'll trace the rise and fall of piecework at a high level,
identifying key figures and ideas during this time.
This section is not intended to be comprehensive:
instead, it sets up the scaffolding necessary for
our later investigations of on--demand work's three questions:
complexity limits,
task decomposition, and
worker relationships.

\subsection{What is piecework? A Primer and Timeline}\label{sec:whatIsPiecework}

\topic{Aligning on--demand work with piecework requires an understanding of what piecework is.}
While it has had several definitions over the years,
we can trace a constellation of characteristics that recur throughout the literature.
We'll follow this research, collecting
descriptions,
examples, and
definitions,
tracing an outline of
\textit{what piecework is}
alongside piecework's contemporary developments in practice.

@@

\topic{We see this sense of independence in myriad times, locales, and industries.}
\citeauthor{10.2307/3827491} offers a look into the lives and culture of ``match--girls''
--- young pieceworkers, mostly women, who assembled matchsticks in the late \nth{19} century.
Of interest was their reputation ``\dots~for generosity, independence, and protectiveness,
but also for brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
\citeauthor{10.2307/27508091} document piecework from 1850 through 1930 in Australia,
finding similar notions of independence and autonomy among piecework newspaper compositors:
``If a piece--work compositor~\dots~decided that he did not want to work on a particular day or night,
the management recognised his right to put a `substitute' or `grass' compositor in his place''~\cite{10.2307/27508091}.
This sense of independence and autonomy appears to be an inherent component of piecework.
%that resonates across decades, industries, and locales where piecework is found.

\ali{I think now that this is moved to this area there's a good opportunity to frame this more as
``now that workers could choose their own schedule, style, etc\dots,
newfound interest in how to manage these workers emerged''.
Thoughts?}
\topic{The early growth of piecework led to discussion on how best to manage pieceworkers,
generally regarding workers antagonistically~\cite{norton1900textile,clark1908cotton}.}
This was a far cry from the earlier rhetoric on piecework, which promised that
piece workers would gladly work as diligently and as hard as possible because
incentive--based pay would reward workers directly for hard work~[].

\topic{Piecework opened the door for people who previously couldn't participate in the labor market to do so, and to acquire job skills incrementally.}
During World War II, women received training in narrow subsets of more comprehensive jobs, enabling work in capacities similar to conventional (i.e. male) workers~\cite{hart2013rise}.
Workers with specific skill subsets could be matched to suitable tasks.
Women previously had virtually no opportunities
to engage in engineering and metalworking apprenticeships as men did;
now, they
could be trained quickly on narrowly scoped tasks,
demonstrate proficiency, and become experts.

\topic{Piecework's popularity in the United States and Europe fell almost as quickly as it had climbed.}
Between 1938 and 1942, the proportion of metal workers under piecework systems had climbed steeply from 11\% to 60\%~\cite{hart2005piecework}.
By 1961, that proportion dropped to 8\%~\cite{carlson1982time}.
\citeauthor{carlson1982time} details that, from 1973 to 1980, the holdouts of piecework
--- where more than 50\% worked under incentive wage plans ---
were principally in clothes--making (e.g. hosiery, footwear, and garments).
\citeauthor{hart2013rise}'s work substantively explores the precipitous decline of piecework in the last third of the \nth{20} century.

@@

\subsection{Case studies in piecework}
\topic{Throughout the rest of the paper, we will return to four case studies to frame our analyses:
Airy's employment of human computers;
domestic and farm work;
the ``match--girls'' strike;
and industrial and assembly-line work.}
In introducing these cases at a high level,
we'll trace the history of piecework
while also framing the later analysis of the major research threads we named earlier: complexity, decomposition, and relationships.

% \onlyinsubfile{\clearpage}
\subsubsection{Airy's Computers}

@@

\msb{This one seems way thinner than the others? Maybe it should be merged with the earlier description of Airy.}
\topic{Some of the first systematic cases of what we would recognize as crowd work
can be found in the study of astronomy.}
In the \nth{19} century, the calculation of celestial bodies had become
a competitive field with 
Airy needed to compute tables that would
allow sailors to locate themselves by starlight from sea.
This work ostensibly called for educated people who comprehensively understood mathematics.
Airy realized that he could break the tasks down and delegate the constituent parts
to ``human computers'' who
``\dots~possessed the basic skills of mathematics,
including `Arithmetic, the use of Logarithms, and Elementary Algebra'~''~\cite{grier2013computers}.
As a result, many of Airy's computers had relatively rudimentary educations
compared to the background of education that typically worked in the calculation of solar tables.
Airy distributed tasks by mail,
allowing work to be completed by a somewhat geographically distributed workforce,
and paid for each piece of work completed.
Airy also instituted a policy of firing his computers once they reached age 23.
%\ali{this has a ``gosh isn't that interesting?'' feeling to it.
%should i just leave it out until later when there's something to say about it
%(namely, that the mode of work was initially designed to stymie professional growth)?}%

The human computers captured several aspects of task decomposition that would become common. First, the work was designed such that it could be done independently and without collaboration. This enabled geographically-distributed workers to complete the tasks. Second, the work was designed so that intermediate results could be quickly verifiable as correct: Airy would have two workers each do the calculation, and a another person compare their answers. Third, Airy identified ways to decompose the large task into narrowly-trainable subtasks.
%\ali{This should go later someplace, but not clear where:

This practice ensured two outcomes that arguably disfavored workers.
First, it eliminated any potential to advance professionally, as
workers' careers in this area ended relatively early in their careers,
and without formal education in mathematics they struggled to find work for which their experience was meaningful.
Second, it limited workers' ability to organize
by ensuring that workers hardly spent sufficient time to successfully rally their peers.

% \onlyinsubfile{\clearpage}
\subsubsection{Domestic and Farmhand Labor}

@@

\topic{The application of piecework to farm work in the late \nth{19} century and
later to manufacturing of small goods, such as garments and matches, at the turn of the \nth{20} century
proved to be a formative period for piecework as we would come to know it.}
Piecework regimes in farms and in homes engaged workers in assembling clothing, where they were paid by the \textit{piece} of output. %, and
%finally in the early sweatshops where women made matchsticks under dangerous and even hazardous conditions.
%\topic{Farm and field work emerged early on as a rich field for the use of piecework,
%perhaps in part due to the relatively low skillset necessary to complete work.}
Textile manufacturers found that they could deliver fabric to people at their homes,
paying them to sew together clothing.
The manufacturers would later return to retrieve the finished garments,
paying these workers for each \textit{piece} of clothing made. 
Farm work applied the idea of piecework by
paying workers for things like picking bushels of fruit or bringing to birth animals~\cite{10.2307/2338394}.
\msb{cutting this next part, it seems irrelevant to the paragraph:}
%\citeauthor{hughRaynbirdTaskWork} argues from another perspective, but along a similar thread, that
%the only factor important in piecework was
%whether the worker's final product was to the ``satisfaction of his master''~\cite{hughRaynbirdTaskWork}.
%Piecework's advantage,
%\citeauthor{hughRaynbirdTaskWork} argues,
%was this hands--off management of workers.

@@

\subsubsection{The Match--Girls' Strike}

@@

\topic{The match--girls strike of 1888 was sparked by a worker's arbitrary docking of pay, but
much deeper resentment had been boiling over for years.}
Match--girls were already frustrated with
the arbitrariness of management,
poor working conditions, and
having to work with phosphoric materials (which was proving hazardous).
\ali{I might need to add some info here, and certainly citations, but there'll be overlap with the other citations here}

@@

%\topic{To understand how these women became such influential, if underreported, figures in labor advocacy's history, we should shine a light onto the reputation that had formed for these workers.}
Match girls were the only group in \citeyear{booth1903life} to have formed a trade union,
according to \citeauthor{booth1903life}'s account at the time~\cite{booth1903life}.
\citeauthor{10.2307/3827491} noted that match girls
``\dots~pooled their resources to purchase their plumes and clothes~\dots~and expressed their solidarity through small [and major] strikes''~\cite{booth1903life}.
But they were also, as \citeauthor{10.2307/3827491} confesses, known for ``brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
These were workers who treasured their independence, but also fiercely protected one another, contributing to the common good.
Their ``brashness'' for instance may have detracted from their public image, but almost undoubtedly contributed to their sense of solidarity,
making their propensity to act against such unfair treatment and poor conditions understandable and maybe predictable.

@@

\subsubsection{Industrial Workers}
\topic{Piecework might be most familiar to the HCI researcher in the context of industrial and factory work, which largely defined manufacturing through the \nth{20} century.}
%In the railroad industry and especially on the assembly line we see many of the mechanisms of work and worker management that later made their way into crowd work.
%Furthermore, it's from here the piecework literature draws most upon to find limits speaking to complexity and task granularity.
%We'll expose these facets of piecework very superficially in the context of industrial work to better speak to these research questions in crowd work later.
\topic{Before the factory assembly line arose, railway companies adopted piecework regimes in the early \nth{20} century. \msb{What was pieceworky about this? They were paid per meter of railroad that they laid down?}
What followed was a flourishing of piecework management practices,
as railway companies worked to find effective ways
to motivate and evaluate this skilled workforce of engineers.}
\citeauthor{10.2307/23702539} takes up a case study of the Santa Fe Railway,
finding that they employed ``efficiency experts'' to develop a ``standard time''
to determine pay for each task at the company informed by
``thousands of individual operations''; % ~\cite{10.2307/23702539}.
\citeauthor{10.2307/23702539} goes on to list
some of the roles required to facilitate piecework
in the early \nth{20} century
--- among them, ``piecework clerks, inspectors, and `experts'~''~\cite{10.2307/23702539}. \topic{This oversight, while controversial
(especially among workers~\cite{american1921problem}),
paved the way for piecework to grow substantially.}

@@

\topic{Piecework %through the \nth{20} century centered around automobile and other mass manufacturing, but
found its way into the war effort during World War II.}
With the vast majority of men drafted into service,
factories found themselves turning to
a predominantly female workforce that had neither
the formal training nor
the years of apprenticeship experience
that conventional workers would have had.
Rather than attempting to train this new labor force in every aspect of industrial work,
these women were trained for individual tasks
% --- such as riveting component to another ---
and assigned to that task. %, decomposed from a more comprehensive task.
``Rosie the Riveter'',
an icon of \nth{20} century America who
represented empowerment and opportunity for women~\cite{honey1985creating},
was a pieceworker~\cite{davies2014origins}.

>>>>

\topic{The HCI community has used the term ``piecework'' to describe
myriad instantiations of on--demand labor, but
researchers have generally made this allusion in passing.}
Since this paper traces a much stronger parallel between
(historical) piecework and (contemporary) on--demand work,
a more comprehensive background on piecework will be useful.
%We will more carefully discuss piecework in this section
%to help make our observations and arguments with better familiarity with the topic.
Specifically,
first, we'll define ``piecework'' as researchers in its field understand it;
and second, we'll trace the rise and fall of piecework at a high level,
identifying key figures and ideas during this time.
This section is not intended to be comprehensive:
instead, it sets up the scaffolding necessary for
our later investigations of on--demand work's three questions:
complexity limits,
task decomposition, and
worker relationships.

\subsection{What is piecework?: A primer and timeline}\label{sec:whatIsPiecework}

\topic{Aligning on--demand work with piecework requires an understanding of what piecework is.}
While it has had several definitions over the years,
we can trace a constellation of characteristics that recur throughout the literature.
We'll follow this research, collecting
descriptions,
examples, and
definitions,
to develop a working understanding.

@@

\topic{We see this sense of independence in myriad times, locales, and industries.}
\citeauthor{10.2307/3827491} offers a look into the lives and culture of ``match--girls''
--- young pieceworkers, mostly women, who assembled matchsticks in the late \nth{19} century.
Of interest was their reputation ``\dots~for generosity, independence, and protectiveness,
but also for brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
\citeauthor{10.2307/27508091} document piecework from 1850 through 1930 in Australia,
finding similar notions of independence and autonomy among piecework newspaper compositors:
``If a piece--work compositor~\dots~decided that he did not want to work on a particular day or night,
the management recognised his right to put a `substitute' or `grass' compositor in his place''~\cite{10.2307/27508091}.
This sense of independence and autonomy appears to be a common component of piecework.
%that resonates across decades, industries, and locales where piecework is found.

\ali{I think now that this is moved to this area there's a good opportunity to frame this more as
``now that workers could choose their own schedule, style, etc\dots,
newfound interest in how to manage these workers emerged''.
Thoughts?}
\topic{Since workers could now choose their own schedule and style, there arose a discussion on how best to manage pieceworkers.}
This conversation generally regarded workers antagonistically~\cite{roy1954efficiency}, a far cry from the earlier rhetoric on piecework, which promised that
piece workers would gladly work as diligently and as hard as possible because incentive--based pay rewards hard work~\cite{clark1908cotton}.

\topic{Piecework opened the door for people who previously couldn't participate in the labor market to do so, and to acquire job skills incrementally.}
During World War II, women received training in narrow subsets of more comprehensive jobs, enabling work in capacities similar to conventional (male) workers~\cite{hart2013rise}.
%Workers with specific skill subsets could be matched to suitable tasks.
Women previously had virtually no opportunities
to engage in engineering and metalworking apprenticeships as men did;
now, they
could be trained quickly on narrowly scoped tasks,
demonstrate proficiency, and become experts.

\topic{Piecework's popularity in the United States and Europe fell almost as quickly as it had climbed.}
Between 1938 and 1942, the proportion of metal workers under piecework systems had climbed steeply from 11\% to 60\%~\cite{hart2005piecework}.
By 1961, that proportion dropped to 8\%~\cite{carlson1982time}.
\citeauthor{carlson1982time} details that, from 1973 to 1980, the holdouts of piecework
--- where more than 50\% worked under incentive wage plans ---
were principally in clothes--making (e.g. hosiery, footwear, and garments).
\citeauthor{hart2013rise} offer a number of explanations for the sudden vanishing of piecework.
The salient suggestions include:
\begin{inlinelist}
\item the emergence of more effective, more nuanced incentive models
--- rewarding teams for complex achievements, for instance;
\item the shifting of piecework industries such as manufacturing and textiles to other countries; and
\item the quality of ``multidimensional'' work becoming too difficult to evaluate
\end{inlinelist}~\cite{hart2013rise}.

@@

\subsection{Case studies in piecework}
\topic{Throughout the rest of the paper, we will return to four case studies to frame our analyses:
Airy's employment of human computers;
domestic and farm workers;
the ``match--girls'' strike;
and industrial and assembly-line workers.}
In introducing these cases at a high level,
we'll trace the history of piecework
while also framing the later analysis of the major research threads we named earlier: complexity, decomposition, and relationships.

% \onlyinsubfile{\clearpage}
\subsubsection{Airy's computers}

@@

\msb{This one seems way thinner than the others? Maybe it should be merged with the earlier description of Airy.}
\topic{In the \nth{19} century, the calculation of celestial bodies had become a competitive field, and Airy needed to compute tables that would allow sailors to locate themselves by starlight from sea.}
This work ostensibly called for educated people who comprehensively understood mathematics.
Airy realized that he could break the tasks down and delegate the constituent parts
to human computers, or people who could compute basic functions.
These human computers ``\dots~possessed the basic skills of mathematics,
including `Arithmetic, the use of Logarithms, and Elementary Algebra'~''~\cite{grier2013computers}.
As a result, many of Airy's computers had relatively rudimentary educations
compared to the background of education that typically worked in the calculation of solar tables.
Airy distributed tasks by mail,
allowing work to be completed by a somewhat geographically distributed workforce,
and paid for each piece of work completed.
Airy also instituted a policy of firing his computers once they reached age 23.
%\ali{this has a ``gosh isn't that interesting?'' feeling to it.
%should i just leave it out until later when there's something to say about it
%(namely, that the mode of work was initially designed to stymie professional growth)?}%

The human computers captured several aspects of task decomposition that would become common. 
First, the work was designed such that it could be done independently and without collaboration. 
%This enabled geographically-distributed workers to complete the tasks. 
Second, the work was designed so that intermediate results could be quickly verified: Airy would have two workers each do the calculation, and another person compare their answers.
Third, Airy identified ways to decompose the large task into narrowly-trainable subtasks.
%\ali{This should go later someplace, but not clear where:

This practice ensured two outcomes that disfavored workers.
First, it eliminated any potential to advance professionally, as
workers' careers ended relatively early in their careers,
and without formal education in mathematics they struggled to find work for which their experience was meaningful.
Second, it limited workers' ability to organize
by ensuring that workers were barely in communication with each other.% hardly spent sufficient time to successfully rally their peers.

% \onlyinsubfile{\clearpage}
\subsubsection{Domestic and farmhand labor}

@@

\topic{The application of piecework to farm work in the late \nth{19} century and
later to manufacturing of small goods, such as garments and matches, at the turn of the \nth{20} century
proved to be a formative period for piecework as we would come to know it.}
Piecework regimes in farms and in homes engaged workers in assembling clothing. %, and
%finally in the early sweatshops where women made matchsticks under dangerous and even hazardous conditions.
%\topic{Farm and field work emerged early on as a rich field for the use of piecework,
%perhaps in part due to the relatively low skillset necessary to complete work.}
Textile manufacturers found that they could deliver fabric to people at their homes, asking them to sew together clothing.
The manufacturers would later return to retrieve the finished garments,
paying these workers for each piece of clothing completed. 
Farm work applied the idea of piecework by
paying workers for tasks like picking bushels of fruit or bringing to birth animals~\cite{10.2307/2338394}.
\msb{cutting this next part, it seems irrelevant to the paragraph:}
%\citeauthor{hughRaynbirdTaskWork} argues from another perspective, but along a similar thread, that
%the only factor important in piecework was
%whether the worker's final product was to the ``satisfaction of his master''~\cite{hughRaynbirdTaskWork}.
%Piecework's advantage,
%\citeauthor{hughRaynbirdTaskWork} argues,
%was this hands--off management of workers.

@@

\subsubsection{The match--girls' strike}

@@

\topic{The match--girls strike of 1888 was sparked by a worker's arbitrary docking of pay, but
much deeper resentment had been simmering for years.}
Match--girls were already frustrated with
the arbitrariness of management,
poor working conditions, and
having to work with hazardous phosphoric materials.
\ali{I might need to add some info here, and certainly citations, but there'll be overlap with the other citations here}

@@

%\topic{To understand how these women became such influential, if underreported, figures in labor advocacy's history, we should shine a light onto the reputation that had formed for these workers.}
Match--girls were the only group in \citeyear{booth1903life} to have formed a trade union,
according to \citeauthor{booth1903life}'s account at the time~\cite{booth1903life}.
\citeauthor{10.2307/3827491} noted that match--girls
``\dots~pooled their resources to purchase their plumes and clothes~\dots~and expressed their solidarity through small [and major] strikes''~\cite{booth1903life}.
But they were also, as \citeauthor{10.2307/3827491} confesses, known for ``brashness, irregularity, low morality, and little education''~\cite{10.2307/3827491}.
These were workers who treasured their independence, but also fiercely protected one another, contributing to the common good.
Their ``brashness'' for instance may have detracted from their public image, but almost undoubtedly contributed to their sense of solidarity,
making their propensity to act against such unfair treatment and poor conditions understandable and maybe predictable.

@@

\subsubsection{Industrial workers}
\topic{Piecework might be most familiar in the context of industrial and factory work, which largely defined manufacturing through the \nth{20} century.}
%In the railroad industry and especially on the assembly line we see many of the mechanisms of work and worker management that later made their way into crowd work.
%Furthermore, it's from here the piecework literature draws most upon to find limits speaking to complexity and task granularity.
%We'll expose these facets of piecework very superficially in the context of industrial work to better speak to these research questions in crowd work later.
\topic{Before the factory assembly line arose, however, railway companies adopted piecework regimes in the early \nth{20} century. \msb{What was pieceworky about this? They were paid per meter of railroad that they laid down?}
What followed was a flourishing of piecework management practices,
as railway companies worked to find effective ways
to motivate and evaluate this skilled workforce of engineers.}
\citeauthor{10.2307/23702539} takes up a case study of the Santa Fe Railway,
finding that they employed ``efficiency experts'' to develop a ``standard time''
to determine pay for each task at the company informed by
``thousands of individual operations''; % ~\cite{10.2307/23702539}.
\citeauthor{10.2307/23702539} goes on to list
some of the roles required to facilitate piecework
in the early \nth{20} century
--- among them, ``piecework clerks, inspectors, and `experts'~''~\cite{10.2307/23702539}. \topic{This oversight, while controversial
(especially among workers~\cite{american1921problem}),
paved the way for piecework to grow substantially.}

@@

\topic{Piecework %through the \nth{20} century centered around automobile and other mass manufacturing, but
found its way into the war effort during World War~II.}
With the vast majority of men drafted into service,
factories found themselves turning to
a predominantly female workforce that had neither
the formal training nor
the years of apprenticeship experience
that conventional workers would have had.
Rather than attempting to train this new labor force in every aspect of industrial work,
these women were trained for individual tasks
% --- such as riveting component to another ---
and assigned to that task. %, decomposed from a more comprehensive task.
``Rosie the Riveter'',
an icon of \nth{20} century America who
represented empowerment and opportunity for women~\cite{honey1985creating},
was a pieceworker~\cite{davies2014origins}.
